---
title: "Assignment 2"
author: "Rakyan Adhikara"
date: '2025-04-11'
output: html_document
---

## Installing libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(GGally)
library(car)
library(stats)
library(tidyverse)
library(readxl)
library(leaps)
library(dplyr)
library(TSA)
library(tseries)
library(astsa)
library(lmtest)
options(scipen = 999)
```

## Reading Dataset

```{r}
df_raw <- read.csv("chart.csv")
head(df_raw)
```

# Data Exploration and Analysis

## Summary of dataset

```{r}
summary(df_raw)
```


### ACF Plot for orginal dataset

```{r, fig.width=10, fig.height=7}
par(mar=c(5,4,4,2),
    cex.main=1,
    cex.lab=1,
    cex.axis=1)

# ACF Plot
acf(df_raw)
```

## Descriptive Analysis

### Converting dataset to time series data

```{r pressure, echo=FALSE}
# Creating the time series data
bitc_ts <- ts(df_raw$Bitcoin,
              frequency = 12, # Set the frequency to 12 since it is monthly
              start = c(2011, 8), # Start on August 2011
              end = c(2025, 2)) # End at January 2025
bitc_ts
```

### Time series plot

```{r, fig.width=10, fig.height=7}
# Time series plot

plot(bitc_ts,
     main = "Time series plot of monthly historical performance of the Bitcoin index (in USD)",
     ylab='US$',
     xlab='Time',
     ylim=c(0,130000000),
     type='o',
     col="#D55E00",
     )
```

### Scatter plot for Time Series data

```{r, fig.width=10, fig.height=7}
plot(y=bitc_ts,
     x=zlag(bitc_ts),
     main = "Scatter plot of monthly historical performance of the Bitcoin index (in USD)",
     ylab='US$',
     xlab='Previous in US$',
     col="#D55E00")
```

### Correlation of the Time Series Dataset

```{r}
# Read the data into bitc_corr_y
bitc_corr_y <- bitc_ts

# Generate first lag of the series into bitc_corr_x
bitc_corr_x <- zlag(bitc_ts) 

# Create an index to get rid of the first NA value in x if applicable
bitc_corr_idx <- 2:length(bitc_corr_x)  

# Calculate correlation between numerical values in x and y
cor(bitc_corr_y[bitc_corr_idx],bitc_corr_x[bitc_corr_idx]) 
```

### Normality Test

```{r fig.width=10, fig.height=7}
# QQ Plot
qqnorm(y = bitc_ts,
       main = "QQ plot of the Bitcoin Price", 
       col = "blue")
qqline(y = bitc_ts, 
       col = 2, 
       lwd = 1, 
       lty = 2)

# Shapiro Test
shapiro.test(bitc_ts)
# p-value = 0.0001 
```


#### ADF Test on Time Series dataset

```{r}
# ADF test on Time Series dataset
adf.test(bitc_ts)
# P-Value = 0.9375 -> Non-Stationary
```

```{r fig.width=10, fig.height=7}
# Create the ACF plot: slowly decay trend
acf(bitc_ts, main = "ACF of Bitcoin Price Performance Time Series")

# convert the ACF plot into a numerical representation
# Calculate ACF values for the Bitcoin Price Performance Time Series
bitc_ts_acf_values <- acf(bitc_ts, plot = FALSE)

# Print the ACF values
bitc_ts_acf_values$acf
```

```{r, fig.width=10, fig.height=7}
# large first lag in the pacf
pacf(bitc_ts, main = "PACF of Bitcoin Price Performance Time Series")

# convert the ACF plot into a numerical representation
# Calculate PACF values for the Bitcoin Price Performance Time Series
bitc_ts_pacf_values <- pacf(bitc_ts, plot = FALSE)

# Print the PACF values
bitc_ts_pacf_values$acf
```

```{r}
summary(bitc_ts)
```

## Data Transformation

### Box-Cox Transformation

```{r, fig.width=10}
# Transformation using yule-walker
bitc_ts_yw <- BoxCox.ar(y = bitc_ts, method = "yule-walker")

title(main = "Log-likelihood versus the values of lambda for Bitcoin Price")

# Values of the first and third vertical lines
bitc_ts_yw$ci

# To find the lambda value of the middle vertical line
bitc_ts_yw_lambda <- bitc_ts_yw$lambda[which(max(bitc_ts_yw$loglike) == bitc_ts_yw$loglike)]
bitc_ts_yw_lambda
```

```{r}
# Apply Box-Cox transformation using the optimal lambda value 
bitc_ts_bc <- ((bitc_ts^bitc_ts_yw_lambda) - 1) / bitc_ts_yw_lambda

# Set the output file and dimensions (width, height) in pixels
png("bitc_ts_bc.png", width = 1200, height = 600)

# Set up the layout for side-by-side plots
par(mfrow = c(1, 2))

# Create a time series plot of the original data
plot(bitc_ts,
     main = "Original Bitcoin Price Performance",
     xlab = "Year",
     ylab = "Price (US$)",
     type = "l",
     col = "black")

# Create a time series plot of the Box-Cox transformed data
plot(bitc_ts_bc,
     main = "Box-Cox transformation Bitcoin Price Performance",
     xlab = "Year",
     ylab = "Box-Cox transformation Price (US$)",
     type = "l",
     col = "blue",
     xlim = c(2011, 2025),
     ylim = c(0, 100))

# Reset the layout
par(mfrow = c(1, 1))
```

#### Normality Evaluation for Box-Cox Transformation

```{r, fig.width=10}
# QQ Plot
qqnorm(y = bitc_ts_bc, 
       main = "QQ plot of Bitcoin Price Performance after Box Cox Transformation", 
       col = "blue")
qqline(y = bitc_ts_bc, 
       col = 2, 
       lwd = 1, 
       lty = 2)

# Shapiro Test
shapiro.test(bitc_ts_bc)
```

#### Stationary Evaluation for Box-Cox Transformation

```{r}
# ADF Test
adf.test(bitc_ts_bc, alternative = "stationary")

# PP Test
pp.test(bitc_ts_bc)

# KPSS Test
kpss.test(bitc_ts_bc, null = "Level")
```

### Log-Difference Transfomration

```{r}
bitc_ts_log <- log(bitc_ts)
bitc_ts_logdiff <- diff(bitc_ts_log)

# Set the output file and dimensions (width, height) in pixels
png("bitc_ts_logdiff.png", width = 1200, height = 600)

# Set up the layout for side-by-side plots

par(mfrow = c(1, 2))
# Create a time series plot of the original data
plot(bitc_ts,
     main = "Original Bitcoin Price Performance Time Series",
     xlab = "Year",
     ylab = "Price (US$)",
     type = "l",
     col = "black")

# Create a time series plot of the log differenced data
plot(bitc_ts_logdiff,
     main = "Log Differenced Bitcoin Price Performance Time Series",
     xlab = "Year",
     ylab = "Differenced Price (US$)",
     type = "l",
     col = "blue")
```

```{r, fig.width=10}
plot(bitc_ts_logdiff, main = "Differenced Log Bitcoin Time Series", ylab = "Differenced log(Bitcoin)", col = "darkgreen")
```

#### Normality Evaluation for Log-Difference Transformation

```{r}
# Shapiro Test
shapiro.test(bitc_ts_logdiff)
```

#### Stationary Evaluation for Log-Difference Transformation

```{r}
# ADF Test
adf.test(bitc_ts_logdiff, alternative = "stationary")

# PP Test
pp.test(bitc_ts_logdiff)

# KPSS Test
kpss.test(bitc_ts_logdiff, null = "Level")
```

### First-Difference transfomration

```{r, fig.width=10}  
# First Difference 
bitc_ts_fdiff <- diff(bitc_ts, differences = 1)

# Set the output file and dimensions (width, height) in pixels
png("bitc_ts_fdiff.png", 
    width = 1200, 
    height = 600)

# Set up the layout for side-by-side plots
par(mfrow = c(1, 2))

# Create a time series plot of the original data
plot(bitc_ts,
     main = "Original Bitcoin Price Performance Time Series",
     xlab = "Year",
     ylab = "Price (US$)",
     type = "l",
     col = "black")

# Create a time series plot of the first differenced data
plot(bitc_ts_fdiff,
     main = "First Differenced Bitcoin Price Performance Time Series",
     xlab = "Year",
     ylab = "Differenced Price (US$)",
     type = "l",
     col = "blue")

```

#### Normality Evaluation for First-Difference Transformation

```{r}
shapiro.test(bitc_ts_fdiff)
```

#### Stationary Evaluation for First-Difference Transformation

```{r}
# ADF Test
adf.test(bitc_ts_fdiff, 
         alternative = "stationary")

# PP Test
pp.test(bitc_ts_fdiff)

# KPSS Test
kpss.test(bitc_ts_fdiff, 
          null = "Level")
```

## After evaluating all transformation, Log-Difference has the best transformation for Bitcoin Price Performance

```{r, fig.width=10}
# Plot the ACF of the Log differenced time series
acf(bitc_ts_logdiff, 
    main = "ACF Log Difference")

# Plot the PACF of the Log differenced time series
pacf(bitc_ts_logdiff, 
     main = "PACF Log Difference")

# Models: {ARIMA(1,1,1)), ARIMA(1,1,7), ARIMA(2,1,7), ARIMA(4,1,0), ARIMA(0,1,7)}
```

```{r}
# EACF
eacf(bitc_ts_logdiff, 
     ar.max = 10, 
     ma.max = 10)

# Models: {ARIMA(2,1,2)}
```

```{r}
bic_ols = armasubsets(y = bitc_ts_logdiff,
                      nar = 10,
                      nma = 10,
                      y.name = 'p',
                      ar.method ='ols')
plot(bic_ols)

# Models: {ARIMA(3,1,2)}
```
```{r}
bic_yw = armasubsets(y = bitc_ts_logdiff,
                     nar = 10,
                     nma = 10,
                     y.name ='p',
                     ar.method ='yule-walker')
plot(bic_yw)
# Models: {ARIMA(1,1,7), AARIMA(4,1,0)}
```

```{r}
bic_burg = armasubsets(y = bitc_ts_logdiff,
                       nar = 10,
                       nma = 10, 
                       y.name='p',
                       ar.method='burg')
plot(bic_burg)
# Models: {ARIMA(1,1,7), AARIMA(4,1,0)}
```


```{r}
fit_arima_models <- function(time_series, arima_orders) {
  models <- list()
  for (order in arima_orders) {
    model <- arima(time_series,
                   order = order,
                   method = 'CSS-ML')
    coef_test <- coeftest(model)
    aic_score <- AIC(model)
    bic_score <- BIC(model)
    models[[paste("ARIMA(",
                  paste(order,
                        collapse = ","),
                  ")",
                  sep = "")]] <- list(model = model,
                                      coef_test = coef_test,
                                      AIC = aic_score,
                                      BIC = bic_score)
  }
  return(models)
}

# Define the list of ARIMA models
arima_list <- list(
  c(0,1,7),
  c(1,1,1), 
  c(1,1,7), 
  c(2,1,2), 
  c(2,1,7),
  c(3,1,2), 
  c(4,1,0)
)

bitc_ts_models <- fit_arima_models(bitc_ts, arima_list)

# Accessing the models and their coefficient tests, AIC and BIC scores:
for (model_name in names(bitc_ts_models)) {
  cat("Model:", model_name, "\n")
  cat("Coefficient test:\n")
  print(bitc_ts_models[[model_name]]$coef_test)
  cat("AIC:", bitc_ts_models[[model_name]]$AIC, "\n")
  cat("BIC:", bitc_ts_models[[model_name]]$BIC, "\n\n")
}
```


```{r}
# Function to fit ARIMA models and compute accuracy
calc_arima_acc <- function(data, arima_orders) {
  models <- list()
  accuracy_measures <- list()
  
  for (order in arima_orders) {
    model <- Arima(data,
                   order = order,
                   method = 'CSS-ML')
    models[[paste0("ARIMA(",
                   paste(order,
                         collapse = ","),
                   ")")]] <- model
    accuracy_measures[[paste0("ARIMA(",
                              paste(order,
                                    collapse = ","),
                              ")")]] <- accuracy(model)[1:7]
  }
  
  df_accuracy <- data.frame(do.call(rbind, accuracy_measures))
  colnames(df_accuracy) <- c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE", "ACF1")
  
  return(df_accuracy)
}

# Define the list of ARIMA models
arima_list <- list(
  c(0,1,7),
  c(1,1,1), 
  c(1,1,7), 
  c(2,1,2), 
  c(2,1,7),
  c(3,1,2), 
  c(4,1,0)
)

# Call the function with your data and the list of ARIMA models
acc_res <- calc_arima_acc(bitc_ts, arima_list)
print(acc_res)

```
# We found ARIMA(2,1,2) model with MASE of 0.2234238 (22.34%) and MAPE of 18.62875 (18.62%) is the most promissing one for Diagnostic Checking

```{r, fig.width=10}
# Fit the ARIMA model
arima212_CSSML <- arima(bitc_ts,
                         order = c(2, 1, 2),
                         method = 'CSS-ML')

arima212_CSSML_fit <- rstandard(arima212_CSSML)

# Set up the multi-panel plot layout
par(mfrow = c(2, 2))

# Time series plot of standardized residuals
plot(arima212_CSSML_fit,
     main = "Time series plot of standardized residuals",
     xlab = 'Time',
     ylab = 'Standardized Residuals',
     type = 'o',
     col = "blue")

# Histogram of standardized residuals
hist(arima212_CSSML_fit,
     main = "Histogram of standardized residuals",
     ylab = 'Standardized Residuals',
     col = "blue")

# QQ plot of standardized residuals
qqnorm(arima212_CSSML_fit, 
       main = "QQ plot of standardized residuals.",
       col = "blue")
qqline(arima212_CSSML_fit,
       col = 2,
       lwd = 1,
       lty = 2)

# ACF plot of standardized residuals
acf(arima212_CSSML_fit,
    main = "ACF plot of standardized residuals.")

```

```{r}
Box.test(arima212_CSSML_fit, type = "Ljung-Box")
```

```{r}
shapiro.test(arima212_CSSML_fit)
```