---
title: "Final Assessment"
author: "Rakyan Adhikara"
date: '2025-05-15'
output: html_document
---



```{r setup, include=FALSE}
## Installing libraries
knitr::opts_chunk$set(echo = TRUE)

suppressWarnings({
  suppressPackageStartupMessages({
  # library(ggplot2)
  # library(GGally)
  # library(car)
  # library(stats)
  # library(tidyverse)
  # library(readxl)
  # library(leaps)
  # library(dplyr)
  # library(TSA)
  # library(tseries)
  # library(astsa)
  # library(lmtest)
  
  # library(forecast)
  # library(fUnitRoots)
  library(fGarch)
  library(rugarch)
  # library(tseries)
  
    
  library(LSTS)
  library(TSA)
  library(forecast)
  library(tseries)
  library(lmtest)



  })
})


source("seasonal_acf_pacf.R")


```



```{r}
## Reading Dataset
df_monthly_raw <- read.csv("Monthly_Avg_USD.csv")
head(df_monthly_raw)
```


```{r}
# Data Exploration and Analysis

## Summary of dataset

summary(df_monthly_raw)
```


```{r}
#Missing values
sum(is.na(df_monthly_raw))
```


```{r, fig.width=10, fig.height=7}

### ACF Plot for orginal dataset
#par(mar=c(5,4,4,2),
#    cex.main=1,
#    cex.lab=1,
#    cex.axis=1)

#par(mfrow=c(1,1))

# ACF Plot
acf(df_monthly_raw)
```

# 1. Descriptive Analysis
```{r pressure, echo=FALSE}
# Creating the time series data
gold_ts <- ts(df_monthly_raw$USD,
              frequency = 12, # Set the frequency to 12 since it is monthly
              start = c(1978, 1), # Start on January 1978
              end = c(2025, 4)) # End at January 2025
##gold_ts
```

```{r}
gold_ts_2005 <- window(gold_ts, start=c(2005, 1))
gold_ts_2005
```

```{r}
summary(gold_ts_2005)
```


```{r, fig.width=10, fig.height=7}
# Time series plot

plot(gold_ts_2005,
     main = "Time series plot of monthly Gold Price per Troy Ounce (in USD)",
     ylab='US$',
     xlab='Time',
     ylim=c(0,4000),
     col="#D55E00",
     )
```

```{r}
decomposed <- decompose(gold_ts_2005)  # works with ts objects
plot(decomposed)

```

```{r}
library(seastests)

# Example using QS test (for monthly data)
qs(gold_ts_2005)


```



```{r, fig.width=10, fig.height=7}
### Scatter plot for Time Series data
plot(y=gold_ts_2005,
     x=zlag(gold_ts_2005),
     main = "Scatter plot of monthly Gold Price per Troy Ounce (in USD)",
     ylab='US$',
     xlab='Previous in US$',
     col="#D55E00")
```



```{r}
### Correlation of the Time Series Dataset
# Read the data into bitc_corr_y
gold_corr_y <- gold_ts_2005

# Generate first lag of the series into bitc_corr_x
gold_corr_x <- zlag(gold_ts_2005) 

# Create an index to get rid of the first NA value in x if applicable
gold_corr_idx <- 2:length(gold_corr_x)  

# Calculate correlation between numerical values in x and y
cor(gold_corr_y[gold_corr_idx],gold_corr_x[gold_corr_idx]) 
```
```{r}

acf(gold_ts_2005, lag.max = 1, plot = FALSE)$acf[1]

```



```{r fig.width=10, fig.height=7}


### Normality Test
# QQ Plot
qqnorm(y = gold_ts_2005,
       main = "QQ plot of the Gold Price", 
       col = "blue")
qqline(y = gold_ts_2005, 
       col = 2, 
       lwd = 1, 
       lty = 2)

# Shapiro Test
shapiro.test(gold_ts_2005)
# p-value = 0.0001 
```




```{r}
# ADF test on Time Series dataset
suppressWarnings({adf.test(gold_ts_2005)})
suppressWarnings({kpss.test(gold_ts_2005)})
# P-Value = 0.99 -> Non-Stationary
```

```{r fig.width=10, fig.height=7}
# Create the ACF plot: slowly decay trend
par(mfrow=c(1,2))
acf(gold_ts_2005, main = "ACF of Gold Price Time Series")
pacf(gold_ts_2005, main = "PACF of Gold Price Time Series")

par(mfrow=c(1,1))

```
```{r, fig.width=10}

seasonal_acf(gold_ts_2005, main = "Seasonal ACF of Gold", lag=48)
seasonal_pacf(gold_ts_2005, main = "Seasonal PACF of Gold", lag=48)

```


```{r, fig.width=10, fig.height=7}
# convert the ACF plot into a numerical representation
# Calculate ACF values for the Gold Price Performance Time Series
gold_ts_acf_values <- acf(gold_ts_2005, plot = FALSE)

# Print the ACF values
gold_ts_acf_values$acf
gold_ts_pacf_values <- pacf(gold_ts_2005, plot = FALSE)

# Print the PACF values
#gold_ts_pacf_values$acf
```

# 2. Transformations and Stationarity Testing
```{r}
plot_full_analysis <- function(original, transformed, title, fig_num_start = 1) {
  
  # 1 & 2: Plot original and transformed series
  par(mfrow=c(1,2),
  mar = c(5, 2, 3, 2),   # margins: bottom, left, top, right
  cex.main = 1,          # title font size
  cex.lab = 1,           # axis label font size
  cex.axis = 1           # axis tick label font size
)


  plot(original, 
       type = "l", 
       col = "black",
       main = paste("Original Gold Price Performance"),
       ylab = "Price (US$)", 
       xlab = "Year")
 
  plot(transformed, 
       type = "l", 
       col = "blue",
       main = paste(title),
       ylab = "Value",
       xlab = "Time")
  
  par(mfrow = c(1, 1))
  
    # QQ Plot
  qqnorm(y = transformed, 
         main = paste("QQ plot", title), 
         col = "blue")
  qqline(y = transformed, 
         col = 2, 
         lwd = 1, 
         lty = 2)
  
  # Shapiro Test
  print(shapiro.test(transformed))
  
    par(mfrow=c(1,2),
  mar = c(5, 2, 3, 2),   # margins: bottom, left, top, right
  cex.main = 0.85,          # title font size
  cex.lab = 1,           # axis label font size
  cex.axis = 1           # axis tick label font size
  )
  acf(transformed, main = paste("ACF of", title))
  pacf(transformed, main = paste("PACF of", title))
  par(mfrow = c(1, 1))
  
  

  print(suppressWarnings(adf.test(transformed)))
  print(suppressWarnings(kpss.test(transformed)))
  print(suppressWarnings(pp.test(transformed)))
}

```


### 1. Box Cox Transformation
```{r, fig.width=10}
# Box Cox Transformation using loglikelihood
lambda <- BoxCox.lambda(gold_ts_2005, method = "loglik")
gold_ts_boxcox <- BoxCox(gold_ts_2005, lambda)

plot_full_analysis(gold_ts_2005, gold_ts_boxcox, "Box Cox Transformed Series", fig_num_start = 11)
```


```{r}
lambda
# quite close to zero which is log-transform
```


### 2. Log Transformation

```{r, fig.width=10}
gold_ts_log <- log(gold_ts_2005)
plot_full_analysis(gold_ts_2005, gold_ts_log, "Log Transformated Series", fig_num_start = 15)

```
### 3. First Differencing

```{r, fig.width=10}
gold_ts_diff <- diff(gold_ts_2005, differences=1)
plot_full_analysis(gold_ts_2005, gold_ts_diff, "First Differenced Series", fig_num_start = 18)
```
### 4. Second Differencing

```{r, fig.width=10}
gold_ts_diff2 <- diff(gold_ts_2005, differences=2)
plot_full_analysis(gold_ts_2005, gold_ts_diff2, "Second Differenced Series", fig_num_start = 22)

```

### 5. Box Cox + First Differencing

```{r, fig.width=10}
gold_ts_boxcox_diff1 <- diff(gold_ts_boxcox, differences=1)
plot_full_analysis(gold_ts_2005, gold_ts_boxcox_diff1, "Box Cox + First Differenced Series", fig_num_start = 22)

```
### 6. Log + First Differencing
```{r, fig.width=10}
gold_ts_log_diff1 <- diff(gold_ts_log, differences=1)
plot_full_analysis(gold_ts_2005, gold_ts_log_diff1, "Log + First Differenced Series", fig_num_start = 22)
```

# 3. Model Specification
### 1. ACF and PACF
```{r, fig.width=10}
# Plot the ACF of the Log differenced time series
acf(gold_ts_log_diff1, 
    main = "ACF Log Difference")

# Plot the PACF of the Log differenced time series
pacf(gold_ts_log_diff1, 
     main = "PACF Log Difference")

#ARIMA (2,0,2),
```

### 2. EACF
```{r}
# EACF
eacf(gold_ts_log_diff1, 
     ar.max = 10, 
     ma.max = 10)

#(0,1 0,2 1,1 1,2)
```

### 3. BIC
```{r}
bic_ols = armasubsets(y = gold_ts_log_diff1,
                      nar = 10,
                      nma = 10,
                      y.name = 'p',
                      ar.method ='ols')
plot(bic_ols)

# (1,3), (1,4), (4,3), (4,4)
```


```{r}

bic_yw = armasubsets(y = gold_ts_log_diff1,
                     nar = 10,
                     nma = 10,
                     y.name ='p',
                     ar.method ='yule-walker')
plot(bic_yw)

# (10,7), (10,9)
```

```{r}
bic_burg = armasubsets(y = gold_ts_log_diff1,
                       nar = 10,
                       nma = 10, 
                       y.name='p',
                       ar.method='burg')
plot(bic_burg)

# (10,7 10,9)
```

# 4. Parameter Estimation

From set of possible ARIMA(p, d, q) models
### 1. Fitting Models
```{r}
run_arima_model <- function(data, p, d, q) {
  # Methods to use
  methods <- c("ML", "CSS", "CSS-ML")
  results <- list()  # Initialize an empty list 
  
  # Loop through all methods
  for (method in methods) {
    model_name <- paste0("ARIMA(", p, ",", d, ",", q, ")_", method)
    
    # Fit the ARIMA model for the current p, d, q and method
    fit <- tryCatch({
      Arima(data, order=c(p, d, q), method=method)
    }, error = function(e) {
      NULL  # Skip if model fails
    })
    
    # If the model fits successfully, store the results
    if (!is.null(fit)) {
      coef_test <- coeftest(fit)
      
      # Print only the coefficient test results for each method
      cat("\n--- Running", model_name, "---\n")
      print(coef_test)
      
      # Store the results in the results list
      results[[model_name]] <- list(model = fit, coeftest = coef_test)
    } else {
      cat("\nError: Model failed to fit for", model_name, "\n")
    }
  }
  
  return(results)
}

#ARIMA (0,0,1) (0,0,2) (1,0,1) (1,0,2) (1,0,3) (1,0,4) (4,0,3) (4,0,4) (10,0,7) (10,0,9)
```

### 2. Coeftest
```{r}
#ARIMA (0,0,1)
m.arima.001 <- run_arima_model (gold_ts_log_diff1, 0,0,1)
```

```{r}
#ARIMA (0,0,2)
m.arima.002 <- run_arima_model (gold_ts_log_diff1, 0,0,2)
```

```{r}
#ARIMA (1,0,1)
m.arima.101 <- run_arima_model (gold_ts_log_diff1, 1,0,1)
```

```{r}
#ARIMA (1,0,2)
m.arima.102 <- run_arima_model (gold_ts_log_diff1, 1,0,2)
```

```{r}
#ARIMA (1,0,3)
m.arima.103 <- run_arima_model (gold_ts_log_diff1, 1,0,3)
```
#ARIMA (5,0,3) (5,0,6)

```{r}
#ARIMA (1,0,4)
m.arima.104 <- run_arima_model (gold_ts_log_diff1, 1,0,4)
```


```{r}
#ARIMA (4,0,3)
m.arima.403 <- run_arima_model (gold_ts_log_diff1, 4,0,3)
```


```{r}
#ARIMA (4,0,4)
m.arima.404 <- run_arima_model (gold_ts_log_diff1, 4,0,4)
```

```{r}
#ARIMA (10,0,7)
m.arima.1007 <- run_arima_model (gold_ts_log_diff1, 10,0,7)
```


```{r}
#ARIMA (10,0,9)
m.arima.1009 <- run_arima_model (gold_ts_log_diff1, 10,0,9)
```

# 5. Goodness of Fit

### 1. AIC and BIC 
```{r}
#ARIMA (0,0,1) (0,0,2) (1,0,1) (1,0,2) (1,0,3) (1,0,4) (4,0,3) (4,0,4) (10,0,7) (10,0,9)
m.arima_list <- c(m.arima.001,m.arima.002, m.arima.101, m.arima.102, m.arima.103, m.arima.104, m.arima.403, m.arima.404, m.arima.1007, m.arima.1009)

```

```{r}
get_model_criteria <- function(results_list, criterion = c("AIC", "BIC")) {
  criterion <- match.arg(criterion)

  model_scores <- data.frame(
    Model = character(),
    AIC = numeric(),
    BIC = numeric(),
    stringsAsFactors = FALSE
  )

  for (name in names(results_list)) {
    model_info <- results_list[[name]]
    
    # Check model structure
    if (!is.null(model_info$model) && !inherits(model_info$model, "try-error")) {
      model <- model_info$model
      
      # Try to compute AIC/BIC safely
      score <- tryCatch({
        data.frame(
          Model = name,
          AIC = AIC(model),
          BIC = BIC(model),
          stringsAsFactors = FALSE
        )
      }, error = function(e) {
        # Silently skip model if there's any error (e.g., CSS or failed fit)
        NULL
      })
      
      # Only add successful results
      if (!is.null(score)) {
        model_scores <- rbind(model_scores, score)
      }
    }
  }

  # Sort by selected criterion
  model_scores <- model_scores[order(model_scores[[criterion]]), ]

  return(model_scores)
}

```

```{r}
get_model_criteria(m.arima_list)
```

### 2. RMSE, MAE 

```{r}
get_model_accuracy <- function(results_list, sort.by = "RMSE", descending = FALSE) {
  
  # Check valid sort metric
  valid_metrics <- c("RMSE", "MAE")
  if (!(sort.by %in% valid_metrics)) {
    stop(paste("Invalid sort.by metric. Choose one of:", paste(valid_metrics, collapse = ", ")))
  }
  
  # Set up an empty data frame
  accuracy_scores <- data.frame(
    Model = character(),
    RMSE = numeric(),
    MAE = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (name in names(results_list)) {
    model <- results_list[[name]]$model
    
    if (!is.null(model) && !inherits(model, "try-error")) {
      tryCatch({
        acc <- accuracy(model)
        acc_row <- data.frame(
          Model = name,
          RMSE = acc[1, "RMSE"],
          MAE = acc[1, "MAE"],
          stringsAsFactors = FALSE
        )
        accuracy_scores <- rbind(accuracy_scores, acc_row)
      }, error = function(e) {
        cat("Error with model", name, ":", e$message, "\n")
      })
    } else {
      cat("Warning: Model", name, "is invalid or NULL, skipping...\n")
    }
  }
  
  # Set rownames
  rownames(accuracy_scores) <- accuracy_scores$Model
  accuracy_scores$Model <- NULL
  
  # Sort by chosen metric
  accuracy_scores <- accuracy_scores[order(accuracy_scores[[sort.by]], decreasing = descending), ]
  
  return(accuracy_scores)
}



```

```{r}
get_model_accuracy(m.arima_list)
```

# 6. Residuals Analysis
```{r}
plot_ljung_box_pvalues <- function(residuals, lag_max = 36) {
  pvals <- sapply(1:lag_max, function(lag) {
    Box.test(residuals, lag = lag, type = "Ljung-Box")$p.value
  })
  
  plot(1:lag_max, pvals, type = "p", pch = 1,  # 'pch=1' is hollow circle
       xlab = "Lag", ylab = "p-value",
       main = "Ljung-Box Test p-values",
       ylim = c(0, 1))
  abline(h = 0.05, col = "red", lty = 2, lwd = 2)  # red dashed line for significance
}


```

```{r}
check_residuals_diagnostics <- function(model, lag_max = 36) {
  res <- residuals(model)
  
  # Residual plot
  plot(res, main = "Residuals Time Series", ylab = "Residuals")
  
  # Histogram and Q-Q Plot
  par(mfrow = c(1, 2))
  hist(res, main = "Histogram of Residuals", xlab = "Residuals", col = "lightblue", border = "white")
  qqnorm(res, main = "Q-Q Plot of Residuals")
  qqline(res, col = 2)
  
  # ACF and PACF
  par(mfrow = c(1, 2))
  suppressWarnings(seasonal_acf(res, lag.max = lag_max, main = "Seasonal ACF of Residuals"))
  suppressWarnings(seasonal_pacf(res, lag.max = lag_max, main = "Seasonal PACF of Residuals"))
  
  # Reset layout
  par(mfrow = c(1, 1))
  
  # Ljung-Box p-value plot
  plot_ljung_box_pvalues(res, lag_max)
  
  # Statistical tests
  cat("Shapiro-Wilk Normality Test:\n")
  print(shapiro.test(res))
  
  cat("\nLjung-Box Test (lag =", lag_max, "):\n")
  print(Box.test(res, lag = lag_max, type = "Ljung-Box"))
}

```

```{r}
check_residuals_diagnostics(m.arima.403$`ARIMA(4,0,3)_CSS-ML`$model)
```

```{r}
summary(m.arima.403$`ARIMA(4,0,3)_CSS-ML`$model)
```



# B. SEASONAL MODELLING
```{r}
gold_stl <- stl(gold_ts_2005, s.window = 12)
plot(gold_stl)
```


```{r}

analyze_seasonal_residuals <- function(ts_data, seasonal_order = c(0,1,0), seasonal_period = 12, lag_max = 36) {
  
  model <- Arima(ts_data, order = c(0,0,0),
                 seasonal = list(order = seasonal_order, period = seasonal_period))
  
  res <- residuals(model)
  
  # Define a function to plot all graphs when called
  plot_func <- function(){
    par(mfrow = c(1, 2))
    plot(ts_data, main = "Original Time Series", ylab = "Value", xlab = "Time")
    plot(res, main = "Time Series Plot of Residuals", ylab = "Residuals", xlab = "Time")
    
    suppressWarnings(seasonal_acf(res, lag.max = lag_max, main = "Seasonal ACF of Residuals"))
    suppressWarnings(seasonal_pacf(res, lag.max = lag_max, main = "Seasonal PACF of Residuals"))
    
    par(mfrow = c(1, 1))
  }
  
  
  # Return residuals + plotting function
  return(list(
    results = plot_func,
    res = res
  ))
}


```

### 1. seasonal (0,1,0)
```{r}

m1seasonal.gold =  analyze_seasonal_residuals(gold_ts_log_diff1, c(0,1,0))
m1seasonal.gold$results()
```




Since there is one significant lag in ACF , we set Q to 1 in the seasonal ARIMA, we don't take PACF since it shows a trend

### 2. seasonal(0,1,1)
```{r}
m2seasonal.gold =  analyze_seasonal_residuals(gold_ts_log_diff1, c(0,1,1))
m2seasonal.gold$results()

```

since no immediate significant seasonal lag, we conclude its now stationary

```{r}
suppressWarnings(adf.test(m2seasonal.gold$res))
```

```{r}
eacf(m2seasonal.gold$res)

# (0,1) (0,2), (1,1), (1,2), (2,1), (2,2), (2,3)
```

```{r}
plot(armasubsets(y=m2seasonal.gold$res, nar = 10, nma=10, y.name='p', ar.method = 'ols'))

# (10,7) (10,9)
```

```{r}
plot(armasubsets(y=m2seasonal.gold$res, nar = 10, nma=10, y.name='p', ar.method = 'yule-walker'))

# (10,7) (10,9)
```

```{r}
plot(armasubsets(y=m2seasonal.gold$res, nar = 10, nma=10, y.name='p', ar.method = 'burg'))

# (10,7) (10,9)
```


# 2. Fitting Model

```{r}
run_seasonal_arima_model <- function(data, p, d, q) {
  # Methods to use
  methods <- c("ML", "CSS", "CSS-ML")
  results <- list()  # Initialize an empty list 
  
  # Loop through all methods
  for (method in methods) {
    model_name <- paste0("SARIMA(", p, ",", d, ",", q, ")_ (0,1,1)[12]", method)
    
    # Fit the ARIMA model for the current method with fixed seasonal order
    fit <- tryCatch({
      Arima(
        data, 
        order = c(p, d, q), 
        seasonal = list(order = c(0, 1, 1), period = 12),
        method = method
      )
    }, error = function(e) {
      NULL  # Skip if model fails
    })
    
    # If the model fits successfully, store and print results
    if (!is.null(fit)) {
      coef_test <- coeftest(fit)
      
      # Print only the coefficient test results for each method
      cat("\n--- Running", model_name, "---\n")
      print(coef_test)
      
      # Store the results
      results[[model_name]] <- list(model = fit, coeftest = coef_test)
    } else {
      cat("\nError: Model failed to fit for", model_name, "\n")
    }
  }
  
  return(results)
}

```


```{r}
# (0,1)
mseasonal001.gold = run_seasonal_arima_model(gold_ts_log_diff1, 0, 0, 1)
```


```{r}
# (0,2)
mseasonal002.gold = run_seasonal_arima_model(gold_ts_log_diff1, 0, 0, 2)
```

```{r}
# (1,1)
mseasonal101.gold = run_seasonal_arima_model(gold_ts_log_diff1, 1, 0, 1)
```

```{r}
# (1,2)
mseasonal102.gold = run_seasonal_arima_model(gold_ts_log_diff1, 1, 0, 2)
```

```{r}
# (2,1)
mseasonal201.gold = run_seasonal_arima_model(gold_ts_log_diff1, 2, 0, 1)
```

```{r}
# (2,2)
mseasonal202.gold = run_seasonal_arima_model(gold_ts_log_diff1, 2, 0, 2)
```

```{r}
# (2,3)
mseasonal203.gold = run_seasonal_arima_model(gold_ts_log_diff1, 2, 0, 3)
```

```{r}
# (10,7)
mseasonal1007.gold = run_seasonal_arima_model(gold_ts_log_diff1, 10, 0, 7)
```

```{r}
# (10,9)
mseasonal1009.gold = run_seasonal_arima_model(gold_ts_log_diff1, 10, 0, 9)
```

# 3. Goodness of Fit
```{r}
# (0,1) (0,2), (1,1), (1,2), (2,1), (2,2), (2,3), (10,7), (10,9)
results_seasonal_list <- c(mseasonal001.gold, mseasonal002.gold, mseasonal101.gold, mseasonal102.gold, mseasonal201.gold, mseasonal202.gold, mseasonal203.gold, mseasonal1007.gold, mseasonal1009.gold)
# Get models sorted by BIC
get_model_criteria(results_seasonal_list , criterion = "BIC")
```

```{r}
# Sort by RMSE ascending (default)
get_model_accuracy(results_seasonal_list )
```

# 4. Residual Analysis
```{r}
check_residuals_diagnostics(mseasonal203.gold$`SARIMA(2,0,3)_ (0,1,1)[12]CSS-ML`$model)
```

# C. GARCH MODELLING
```{r}
r.gold_ts_2005 = diff(log(gold_ts_2005))*100
plot(gold_ts_2005, type = "l", col = "blue", main = "Log-Differenced Gold Price", ylab = "Value", xlab = "Time")


```

```{r}
McLeod.Li.test(y=gold_ts_2005, main ="McLeod - Li test statistics for Gold Price")
```

```{r}
adf.test(r.gold_ts_2005)
```

Since we already have our best ARMA model from earlier we use that; ARMA (4,0,3)

```{r}
model403cssml <- arima (r.gold_ts_2005, order= c(4,0,3), method='CSS-ML')
coeftest(model403cssml)
```


```{r}
check_residuals_diagnostics(model403cssml)
```

```{r}
abs.r.res.gold <- abs(residuals(model403cssml))
sqr.r.res.gold <- residuals(model403cssml)^2
```

```{r}
par(mfrow= c(1,2))
acf(abs.r.res.gold, main = "ACF plot for absolute return series.", lag=48)

pacf(abs.r.res.gold, main = "PACF plot for absolute return series.", lag=48)


# max(p,q)= 1 from PACF plot and q = 1
# max(p, q=1)=1, p=0,1

# GARCH(0,1), GARCH(1,1)
```


```{r}
eacf(abs.r.res.gold)

# (0,4), (0,5) (1,4), (1,5), (2,4), (2,5), (3,2), (3,3), (3,4)

# max(p,q) = 0 q=4 => max(p,q=4)=0 does not lead to any model
# max(p,q) = 0 q=5 => max(p,q=5)=0 does not lead to any model
# max(p,q) = 1 q=4 => max(p,q=4)=1 does not lead to any model
# max(p,q) = 1 q=5 => max(p,q=5)=1 does not lead to any model
# max(p,q) = 2 q=4 => max(p,q=4)=2 does not lead to any model
# max(p,q) = 2 q=5 => max(p,q=5)=2 does not lead to any model
# max(p,q) = 3 q=2 => max(p,q=2)=3 => p = 3 => GARCH(3,2)
# max(p,q) = 3 q=3 => max(p,q=3)=3 => p = 0,1,2,3 => GARCH(0,3), GARCH(1,3), GARCH(2,3), GARCH(3,3) 
# max(p,q) = 3 q=4 => max(p,q=4)=3 does not lead to any model

# GARCH(3,2), GARCH(0,3), GARCH(1,3), GARCH(2,3), GARCH(3,3)
```


```{r}
par(mfrow= c(1,2))
acf(sqr.r.res.gold, main = "ACF plot for squared return series.")

pacf(sqr.r.res.gold, main = "PACF plot for squared return series.")

# max(p,q)= 2 from PACF plot and q = 3
# max(p, q=3)=2 => does not lead to any model
```

```{r}
eacf(sqr.r.res.gold)

# (0,4), (0,5), (1,1), (1,2), (2,2), (2,3), (3,3), (3,4)

# max(p,q) = 0 q=4 => max(p,q=4)=0 does not lead to any model
# max(p,q) = 0 q=5 => max(p,q=5)=0 does not lead to any model
# max(p,q) = 1 q=1 => max(p,q=1)=1 => p = 0,1 => GARCH(0,1), GARCH(1,1)
# max(p,q) = 1 q=2 => max(p,q=2)=1 does not lead to any model
# max(p,q) = 2 q=2 => max(p,q=2)=2 => p = 0,1,2 => GARCH(0,2), GARCH(1,2), GARCH(2,2)
# max(p,q) = 2 q=3 => max(p,q=3)=2 does not lead to any model
# max(p,q) = 3 q=3 => max(p,q=3)=3 => p = 0,1,2,3 => GARCH(0,3), GARCH(1,3), GARCH(2,3), GARCH(3,3)
# max(p,q) = 3 q=4 => max(p,q=4)=3 does not lead to any model

# GARCH(0,1), GARCH(1,1), GARCH(0,2), GARCH(1,2), GARCH(2,2), GARCH(0,3), GARCH(1,3), GARCH(2,3), GARCH(3,3)
```


```{r}
# GARCH(0,1)
model_403_01 <- fGarch::garchFit(~arma(5,3)+garch(1,0),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_01)
```


```{r}
# GARCH(1,1)
model_403_11 <- fGarch::garchFit(~arma(4,3)+garch(1,1),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_11)
```


```{r}
# GARCH(0,2)
model_403_02 <- fGarch::garchFit(~arma(4,3)+garch(2,0),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_02)

```

```{r}
# GARCH(1,2)
model_403_12 <- fGarch::garchFit(~arma(4,3)+garch(2,1),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_12)
```

```{r}
# GARCH(2,2)
model_403_22 <- fGarch::garchFit(~arma(4,3)+garch(2,2),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_22)
```

```{r}
# GARCH(0,3)
model_403_03 <- fGarch::garchFit(~arma(4,3)+garch(3,0),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_03)
```

```{r}
# GARCH(1,3)
model_403_13 <- fGarch::garchFit(~arma(4,3)+garch(3,1),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_13)
```

```{r}
# GARCH(2,3)
model_403_23 <- fGarch::garchFit(~arma(4,3)+garch(3,2),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_23)
```


```{r}
# GARCH(3,2)
model_403_32 <- fGarch::garchFit(~arma(4,3)+garch(2,3),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_32)
```

```{r}
# GARCH(3,3)
model_403_33 <- fGarch::garchFit(~arma(4,3)+garch(3,3),
                                 data=r.gold_ts_2005, trace=F)
summary(model_403_33)
```

```{r}
# GARCH(0,1), GARCH(1,1), GARCH(0,2), GARCH(1,2), GARCH(2,2), GARCH(0,3), GARCH(1,3), GARCH(2,3), GARCH(3,2), GARCH(3,3)
df <- data.frame(
  Model = c("ARMA(4,3) + GARCH(0,1)",
            "ARMA(4,3) + GARCH(1,1)",
            "ARMA(4,3) + GARCH(0,2)",
            "ARMA(4,3) + GARCH(1,2)",
            "ARMA(4,3) + GARCH(2,2)",
            "ARMA(4,3) + GARCH(0,3)",
            "ARMA(4,3) + GARCH(1,3)",
            "ARMA(4,3) + GARCH(2,3)",
            "ARMA(4,3) + GARCH(3,2)",
            "ARMA(4,3) + GARCH(3,3)"),
  AIC = c(model_403_01@fit$ics[1],
          model_403_11@fit$ics[1],
          model_403_02@fit$ics[1],
          model_403_12@fit$ics[1],
          model_403_22@fit$ics[1],
          model_403_03@fit$ics[1],
          model_403_13@fit$ics[1],
          model_403_23@fit$ics[1],
          model_403_32@fit$ics[1],
          model_403_33@fit$ics[1]),
  BIC = c(model_403_01@fit$ics[2],
          model_403_11@fit$ics[2],
          model_403_02@fit$ics[2],
          model_403_12@fit$ics[2],
          model_403_22@fit$ics[2],
          model_403_03@fit$ics[2],
          model_403_13@fit$ics[2],
          model_403_23@fit$ics[2],
          model_403_32@fit$ics[2],
          model_403_33@fit$ics[2])
)

df <- df[order(df$BIC, decreasing = FALSE), ]
print(df)


```

## Residual analysis
```{r}
check_residuals_diagnostics_fgarch <- function(model, lag_max = 36) {
  # Extract standardized residuals from fGarch model
  res <- residuals(model, standardize = TRUE)
  
  # Plot residuals time series
  plot(res, type = "l", main = "Residuals Time Series", ylab = "Standardized Residuals")

  
  # Histogram and Q-Q plot
  par(mfrow = c(1, 2))
  hist(res, main = "Histogram of Residuals", xlab = "Residuals", col = "lightblue", border = "white")
  qqnorm(res, main = "Q-Q Plot of Residuals")
  qqline(res, col = 2)
  
  # ACF and PACF plots
  par(mfrow = c(1, 2))
  suppressWarnings(acf(res, lag.max = lag_max, main = "ACF of Residuals"))
  suppressWarnings(pacf(res, lag.max = lag_max, main = "PACF of Residuals"))
  
  # Reset layout
  par(mfrow = c(1, 1))
  
  # Plot Ljung-Box test p-values using your custom function
  plot_ljung_box_pvalues(res, lag_max)
  
  # Statistical tests output
  cat("Shapiro-Wilk Normality Test:\n")
  print(shapiro.test(res))
  
  cat("\nLjung-Box Test (lag =", lag_max, "):\n")
  print(Box.test(res, lag = lag_max, type = "Ljung-Box"))
}

```

```{r}
check_residuals_diagnostics_fgarch(model_403_12)
```

## Compare all best models

```{r}
arima_css_ml <- Arima(gold_ts_log_diff1, order = c(4,0,3), method = "CSS-ML")
summary(arima_css_ml)
```


```{r}
library(forecast)
library(fGarch)

# Fit models
arima_css_ml <- Arima(gold_ts_log_diff1, order = c(4,0,3), method = "CSS-ML")
sarima_cssml <- Arima(gold_ts_log_diff1, 
                      order = c(2,0,3), 
                      seasonal = list(order = c(0,1,1), period = 12),
                      method = "CSS-ML")
garch_model <- garchFit(~ arma(4, 3) + garch(2, 1), data = gold_ts_log_diff1, trace = FALSE)


acc_arima_ml <- accuracy(fitted(arima_css_ml), gold_ts_log_diff1)[1,]
acc_sarima <- accuracy(fitted(sarima_cssml), gold_ts_log_diff1)[1, ]
acc_garch <- accuracy(fitted(garch_model), gold_ts_log_diff1)[1, ]

#AIC = c(AIC(arima_css_ml), AIC(sarima_cssml), garch_model@fit$ics["AIC"])

results <- data.frame(
  Model = c("ARIMA(4,0,3)_CSS_ML", "SARIMA(2,0,3)(0,1,1)[12]_CSS-ML", "GARCH(1,2)+ARMA(4,3)"),
  AIC = c(AIC(arima_css_ml), AIC(sarima_cssml), garch_model@fit$ics["AIC"]),
  BIC = c(BIC(arima_css_ml), BIC(sarima_cssml), garch_model@fit$ics["BIC"]),
  RMSE = c(acc_arima_ml["RMSE"], acc_sarima["RMSE"], acc_garch["RMSE"]),
  MAE = c(acc_arima_ml["MAE"], acc_sarima["MAE"], acc_garch["MAE"])
)

print(results)

```


# D. FORECASTING

```{r}
plot_forecast_original_scale <- function(original_ts, log_diff_model, h = 10) {
  require(forecast)

  # 1. Forecast
  fc <- forecast(log_diff_model, h = h)

  # 2. Get last log price
  last_log_price <- log(tail(original_ts, 1))

  # 3. Forecast in log-level
  log_fc_mean <- cumsum(as.numeric(fc$mean)) + as.numeric(last_log_price)
  log_fc_lower <- cumsum(as.numeric(fc$lower[, 2])) + as.numeric(last_log_price)
  log_fc_upper <- cumsum(as.numeric(fc$upper[, 2])) + as.numeric(last_log_price)

  # 4. Convert to original price scale
  price_fc_mean <- exp(log_fc_mean)
  price_fc_lower <- exp(log_fc_lower)
  price_fc_upper <- exp(log_fc_upper)

  # 5. Time setup
  freq <- frequency(original_ts)
  end_time <- end(original_ts)

  fc_time <- seq(end_time[1] + (end_time[2] / freq),
                 by = 1 / freq,
                 length.out = h)

  ts_price_fc_mean <- ts(price_fc_mean, start = fc_time[1], frequency = freq)
  ts_price_fc_lower <- ts(price_fc_lower, start = fc_time[1], frequency = freq)
  ts_price_fc_upper <- ts(price_fc_upper, start = fc_time[1], frequency = freq)

  # 6. Plot
  ts.plot(original_ts, ts_price_fc_mean, ts_price_fc_lower, ts_price_fc_upper,
          col = c("black", "blue", "red", "red"),
          lty = c(1, 1, 2, 2), lwd = c(1, 2, 1, 1),
          main = "Forecast on Original Price Scale",
          ylab = "Price")

  legend("topleft",
         legend = c("Original", "Forecast", "95% CI"),
         col = c("black", "blue", "red"),
         lty = c(1, 1, 2), lwd = c(1, 2, 1), bty = "n")
}

```

```{r}
# Fit ARIMA using CSS method (or replace with "ML" or "CSS-ML" if preferred)
m1.arima.gold_ts <- Arima(gold_ts_log_diff1, order = c(4, 0, 3), method = "CSS-ML")

# Plot forecast in original price scale
plot_forecast_original_scale(gold_ts_2005, m1.arima.gold_ts, h = 10)


```

```{r}
library(forecast)

# Helper to subset time series from a start year (adjust if monthly/quarterly)
subset_ts_from_year <- function(ts_data, start_year) {
  start_time <- start(ts_data)
  freq <- frequency(ts_data)
  
  # Extract time info
  start_year_ts <- floor(start_time[1])  # year part
  start_period <- start_time[2]          # sub-year period (month, quarter)
  
  # Calculate index for new start
  new_start <- c(start_year, 1)
  
  window(ts_data, start = new_start)
}

```



```{r}
years <- c(2010, 2020, 2024)
h <- 10  # forecast horizon

for (yr in years) {
  subset_data <- subset_ts_from_year(gold_ts, yr)
  
  # Log and difference transform on subset
  gold_ts_log <- log(subset_data)
  gold_ts_log_diff1 <- diff(gold_ts_log, differences = 1)
  
  # Fit ARIMA with CSS-ML method
  model_arima <- Arima(gold_ts_log_diff1, order = c(4,0,2), method = "CSS-ML")
  
  # Plot forecast on original scale
  plot_forecast_original_scale(subset_data, model_arima, h = h)
}

```

